Deriving an Executable Model of a Data Stream Protocol from its Specification

### 2009-09-17 papers/ModelBasedTestCaseGenerationDataStreamProtocol/paper.tex
paper ModelBasedTestCaseGenerationDataStreamProtocol: specification initial description

  A custom medium-level Java driver creates and parses the \STSB streaming 
  communication packets. 
  This latter Java driver is formally specified, validated with rigorous
  unit testing, and verified using extended static checking. 
  Because of this process and combination of technologies, using this
  specification we are able to both communicate with a real \ST
  \textbf{and} generate a formally specified, executable, functional
  model-based specification for the protocol itself written in the
  Java Modeling Language.
  \todo executable? how? what does this mean?
  This protocol model is itself an executable
  stream simulator that can be used to develop new \ST user
  applications without having the \ST hardware at all. 
  Consequently,
  from this single specification, we can test that the \ST hardware
  communicates as expected, test that our \ST simulator communicates
  as expected, and simulate the \ST protocol stream itself.

@@

Because of hard time constraints, and the initial unavailability of the 
custom board, we needed to progress concurrently with all the development tasks: 
the specification of the communication protocol with the /SB, the development of 
the controlling code in the /SB, the development of the driver to communicate 
with the /SB, and the development of the simulators needed for future tests 
of the system (the controlling code of the board was developed by the external 
manufacturer of the board, hence, it is not taken into account here).

@@

To tackle the complexity of the concurrent tasks the formal model of the protocol 
was used as a focus point that lead all the other tasks. 
The formal model of the protocol was used to check at run time the correctness of 
the /SB, the driver, the simulator, and the hand made test suites of the driver 
and the simulator.
The formal model was used to automatically generate a test suite to verify that
both the driver and the simulator were compliant to the specification, and eventually 
to verify the specification itself.

@@

In section \ref{sec:data_stream_and_protocol_model} the data stream model 
and protocol is presented, we will present both a high level overview of the 
protocol, and a formalized, JML based, specification.

@@

Finally in section \ref{sec:conclusions} conclusions are drawed and 
ongoing and future works on similar topics is introduced.

%=====================================================================
\section{Data stream and protocol model}
\label{sec:data_stream_and_protocol_model}

@@

The protocol is an asynchronous fixed length packed based protocol.
We will only examine the output sensor data packet, since it is the more used 
and interesting one.

The internal structure of the output sensor data packet reflects the board 
capabilities and sensors.
Basically a single packet has to accommodate various types of data:
\begin{itemize}
 \item Fast data rate streams: audio data, 48KHz or 96KHz.
 \item Medium data rate streams: 5KHz.
 \item Slow data rate streams: temperature, pression, accelerometer, etc.; all 
 these sensor have a data rate far lower than 5 KHz.
 \item Metadatas: metadatas on sensors and board state.
\end{itemize}

Even if we have to deal with data streams that have different data rates, the 
internal structure of the package (as we will see) is strictly fixed. 
This choice originates from a trade-off between two constraints: the efficiency 
of the transmission (redundant data to be minimized) and the complexity to build 
a package (complexity to be minimized).
The complexity minimization constraint is simple to understand: it is the FPGA 
on the \STSB that build the packet, but it has also to be used for various 
other specific tasks. 
Hence, it is important to maintain the building packet code on the FPGA as 
simple and short as possible, to leave enough space for the code implementing 
other tasks.
The Output sensor data packet structure is showed in Table 
\ref{tab:data_packet_structure}, the packet total length is 1024 bytes (512 16 
bit words).

\begin{table}[htbp]
\caption{Output sensor data packet structure.}
\label{tab:data_packet_structure}
\begin{center}
\begin{tabular}{|c|l|}\hline
\textbf{\textit{index}} & \textbf{\textit{function}}\\\hline
0-1 & sentinel\\\hline
2-7 & slow data\\\hline
8-11 & metadata: board state\\\hline
12-13 & reserved\\\hline
14-15 & metadata: board state\\\hline
16 & reserved\\\hline
17-22 & subsequence 1\\\hline
\ldots & \ldots\\\hline
503-508 & subsequence 82\\\hline
\end{tabular}
\end{center}
\end{table}

The subsequence accommodate data from the fast and medium data rate streams 
(there are 4 fast data rate stream and 8 medium data rate stream channels). 
There are 82 subsequences in a packet.
The subsequence structure is shown in Table \ref{tab:subsequence_structure}.

@@

Looking at the structure described and at the data rates frequencies, it
can be seen that there must be something subtle, to make all the figures to 
add up.
The subtleties lies in two forms of redundancies, for slow and medium data 
rates:

@@

\todo describe high level representation of package: sensors and data 
details remains, exact package structure and sensor data location in the
package are lost

\todo describe JML specification of the high level representation of the 
package

@@

@@

\todo benefit: the approach can be used to test different part of the system
that share the same specification

\todo testing the stream

\todo testing the simulator

>>>>

  A custom medium-level Java driver creates and parses the \STSB streaming 
  communication packets. 
  The driver is formally specified, validated and verified with rigorous
  unit testing. 
  Because of this process and combination of technologies, using this
  specification we are able to both communicate with a real \ST
  \textbf{and} generate a formally specified, executable, functional
  model-based specification for the protocol itself written in the
  Java Modeling Language.
  \notev{executable? how? what does this mean?}
  This protocol model is itself an executable
  stream simulator that can be used to develop new \ST user
  applications without having the \ST hardware at all. 
  Consequently,
  from this single specification, we can test that the \ST hardware
  communicates as expected, test that our \ST simulator communicates
  as expected, and simulate the \ST protocol stream itself.

@@

Because of hard time constraints, and the initial unavailability of the 
custom board, we needed to progress concurrently with all the development tasks: 
the specification of the communication protocol with the \STSB, the development of 
the controlling code in the \STSB, the development of the driver to communicate 
with the \STSB, and the development of the simulators needed for future tests 
of the system (the controlling code of the board was developed by the external 
manufacturer of the board, hence, it is not taken into account here).

@@

To tackle the complexity of the concurrent tasks the formal model of the protocol 
was used as a focus point that led all the other development tasks. 
The formal model of the protocol was used to check at run time the correctness of 
the /SB, the driver, the simulator, and the hand made test suites of the driver 
and the simulator.
The formal model was used to automatically generate a test suite to verify that
both the driver and the simulator were compliant to the specification, and eventually 
to verify the specification itself.

@@

In section \ref{sec:data_stream_and_protocol_description} the data stream model 
and protocol is presented.

In section \ref{sec:data_stream_and_protocol_specification} the data stream model 
and protocol is specified.

@@

Finally in section \ref{sec:conclusion} conclusions are drawed and 
ongoing and future works on similar topics is introduced.

%=====================================================================
\section{Data stream and protocol description}
\label{sec:data_stream_and_protocol_description}

@@

The protocol is an asynchronous fixed length packed based protocol.
The protocol has been divided in multiple layers to ease its formal description.
The rationale is the following: mixing up different abstraction levels of a 
protocol description can lead to overly complex and unmanageable descriptions, 
expecially when dealing with formal ones.
The layers identified in protocol are the following:

\begin{description}
 \item[Packet byte structure]: the internal representation of the packet.
 \item[Packet info structure]: the meaningful fields contained in the packets.
 \item[Single packet rules]: the content acceptable values, and how they 
influence each other.
 \item[Packet sequence rules]: the content acceptable values, based on values 
of previously received packets.
 \item[Packet input output asynchronous rules]: the content acceptable values 
and reaction constraints on output packets, based on previously transmitted 
input packets.
\end{description}

We will only examine the output sensor data packet, since it is the more used 
and interesting one.
We will examine the first three layers, that is: the packet byte structure, the 
packet info structure and the single packet rules.

The packet byte and info structures of the output sensor data packet reflects the board 
capabilities and sensors.
Basically a single packet has to accommodate various types of data:
\begin{itemize}
 \item Fast data rate streams: audio data, 48KHz or 96KHz.
 \item Medium data rate streams: 5KHz.
 \item Slow data rate streams: temperature, pression, accelerometer, etc.; all 
 these sensor have a data rate far lower than 5 KHz.
 \item Metadatas: metadatas on sensors and board state.
\end{itemize}

Even if we have to deal with data streams that have different data rates, the 
internal structure of the package (as we will see) is strictly fixed. 
This choice originates from a trade-off between two constraints: the efficiency 
of the transmission (redundant data to be minimized) and the complexity to build 
a package (complexity to be minimized).
The complexity minimization constraint rationale is the following: the FPGA 
on the \STSB builds the packet, but the FPGA is also to be used for various 
other specific tasks; it is important to maintain the building packet code on the 
FPGA as simple and short as possible, to leave enough space to implement other 
tasks.
The Output sensor data packet structure is showed in Table 
\ref{tab:data_packet_structure}, the packet total length is 1024 bytes (512 16 
bit words).

\begin{table}[htbp]
\caption{Output sensor data packet structure.}
\label{tab:data_packet_structure}
\begin{center}
\begin{tabular}{|c|l|}\hline
\textbf{\textit{index}} & \textbf{\textit{function}}\\\hline
0-1 & sentinel\\\hline
2-7 & slow data\\\hline
8-11 & metadata: board state\\\hline
12-13 & reserved\\\hline
14-15 & metadata: board state\\\hline
16 & reserved\\\hline
17-22 & frame 1\\\hline
\ldots & \ldots\\\hline
503-508 & subsequence 82\\\hline
\end{tabular}
\end{center}
\end{table}

The frames accommodate data from the fast and medium data rate streams 
(there are 4 fast data rate stream and 8 medium data rate stream channels). 
There are 82 frames in a packet.
The frame structure is shown in Table \ref{tab:subsequence_structure}.

@@

Looking at the structure described and at the data rates frequencies, it
can be seen that there must be something subtle, to make all the figures to 
add up, that is, to be capable to accommodate three different rates outputs
in a fixed structure, fixed length packet.
The subtleties lies in two forms of redundancies, for slow and medium data 
rates:

@@

The single packet rules delimit the boundaries of the values obtainable from 
the packet: each defined sensor represented in the packet, as well as the 
metadata describing the \STSB and the packet and frame contents,  have a defined 
range of acceptable values.
There are also rules affecting more than one value, basically validity rules (i.e. if 
a specific sensor is active than the reading value has to be in a defined range), 
and rules specifing a correct sequence of frames (i.e. there can be only one switch 
in sampling rate in the frame sequence of a single packet).

%=====================================================================
\section{Data stream and protocol specification}
\label{sec:data_stream_and_protocol_specification}

The specification of the protocol has been used as a focus point to lead all the 
other development tasks.

We needed a specification capable of:

\begin{itemize}
 \item Be executed at run time. If something on the protocol is not respected, 
the run time execution environment is capable to detect and signal the fault.
 \item Generate tests. 
 \item Verify simulators.
 \item Be used in conjunction of hand written unit tests.
\end{itemize}

We also needed a specification capable to tackle properly all the protocol layers.
Since the different layers have very different characteristics, different context
dependant specification approaches were chosen.

\paragraph{Packet byte structure specification}

The specification is provided by a reference Java implementation capable of 
recognizing the proper packet structure in a byte data stream. The implementation 
is tricky, since it checks for the sentinels repeatedly to be sufficiently confidant
that the byte stream can be properly decomposed into a packet stream.

The implementation can be obviously executed at run time (the specification is the 
implementation code). 
Meaningful automatic tests cannot be generated. 
The implementation can be used to verify a packet byte structure simulator. 
The implementation were actually driven (Test Driven Development) by the test suite.

\paragraph{Packet info structure specification}

The specification is provided both by the Java API that has been built, and by the 
annotation with JML language\footnote{
\myhref{http://www.cs.ucf.edu/~leavens/JML/}{The Java Modeling Language (JML)}}. 
JML has been chosen because it is capable to specify Java code: JML is a behavioral 
interface specification language. 
JML specification language combines the design by contract approach (applayed to 
Java language) and the model-based specification approach of interface specification 
languages.

\todo insert proper JML references

Many tools are available supporting JML specification language, focusing on different 
aspects of the language. 
JML2 suite of tools
\footnote{\myhref{http://sourceforge.net/projects/jmlspecs/}{Common JML tools}}
was chosen, since both run time verification support and test generation support are 
provided directly by the suite.

\todo insert proper JML2 references

The Java API specification part is checked automatically by the Java compiler, while
the JML specification part can be checked at run time using the provided JML2 run time 
environment. 
Meaningful automatic tests can be generated from the JML specification. 
The JML specification can be used to verify an abstract packet generation simulator, 
making the simulator an extension of the Java interfaces. 
The specification can also be used to verify the real driver implementation, based on 
the packet byte structure implementation presented above. 
The implementation was driven by the JML specification and by an hand made test suite.
Finally, hand made unit tests can be checked, during execution, against the 
specification, using the provided JML2 run time environment.

\paragraph{Single packet rules}

The specification is provided exclusively by annotations with JML language.

The capabilities of a JML specification have been already descibed in the previous 
paragraph, that is: specification checkable at run time, tests generation, simulator 
verification, usable with hand written tests.

\subsection{On simulators}
\label{subsec:on_simulators}

\todo different layers need different simulators

\todo describe simulators structure

\subsection{A JML specification example}
\label{subsec:a_jml_specification_example}

\todo find interesting JML specification

\todo describe the example

@@

\notev{maybe this part is to be saved for a different paper? BON has 
not been used properly}

@@

\todo benefit: the approach can be used to test different part of the system
that share the same specifications

\todo testing the protocols layers

\todo testing the simulators
